The Role of Variability in Learning Generalization: A Computational Modeling Approach

### 2024-05-28 Presentation/_hold_slides.qmd
update slides

------------------------------------------------------------------------

>>>>

---------------------------------------------------------

---------------

## Project 2 - Experiment 2 Results

::: {.columns}
::: {.column width="50%"}
![](assets/htw_e2_test1.png){width="75%"}
![](assets/htw_e2_test3.png){width="75%"}
:::

::: {.column width="50%"}
![](assets/htw_e2_test2.png){width="75%"}
![](assets/htw_e2_test4.png){width="75%"}
:::
:::

## Project 2 - Experiment 2 Results {.smaller margin="0"}

::: {.columns}
::: {.column width="50%"}
![](assets/htw_e2_test1.png){width=65%"}
:::
::: {.column width="50%"}
![](assets/htw_e2_test2.png){width="65%"}
:::
:::

::: fragment
::: {.columns}
::: {.column width="50%"}

![](assets/htw_e2_test3.png){width="65%"}
:::
::: {.column width="50%"}
![](assets/htw_e2_test4.png){width="65%"}
:::
:::
:::

### 2024-05-28 Presentation/slides.qmd
update slides

---
title: "The Role of Variability in Learning Generalization: A Computational Modeling Approach"
page-layout: full
aliases:
  - /slides.html
css: /Assets/Style/slides.css
title-slide-attributes:
  data-background-image: /Presentation/assets/sol_blank.png
  data-background-size: 300px #contain
  data-background-opacity: "0.15"
  data-background-position: "0% 0%"
format-links: [clean-revealjs, pptx,gfm]
format: 
  html: 
    subtitle: "Dissertation Defense"
  clean-revealjs:
    #theme: moon #white
    #output-ext: revealjs.html
    # template-partials:
    #   - assets/title-slide.html
    output-file: slides_revealjs
    date: 05/28/204
    slideNumber: true
    hash-type: number # Use slide numbers in the URL hash
    center: true
    scrollable: true
    chalkboard:
      theme: whiteboard
  # pptx:
  #   incremental: false
  # hugo-md:
  #   include: false
  #   html-math-method: mathjax
  #   output-file: slides_hugo.md
  # gfm:
  #   output: false
  #   output-file: slides_gfm.md
  # beamer:
  #   incremental: false
  #   colortheme: lily 
author:
  - name: Thomas Gorman
    url: https://tegorman13.github.io
    orcid: 0000-0001-5366-5442
    affiliations:
      - name: Indiana University
        department: Psychological & Brain Sciences, Cognitive Science
# params:
#   start_slide: 
#engine: knitr
---

@@

## Issues with previous research

@@

:::: columns
::: {.column}
![](assets/kerr_booth.png)
:::
::: {.column}
- bean bag throwing task
- Constant and varied conditions train from distinct positions
- Both groups are tested from the position where the constant group trained
- Impressive demonstration of varied training outperforming constant training 
:::
::::
::: footer
@kerrSpecificVariedPractice1978
:::

@@

- instances refers to individual throws (x and y velocity)
- similarity refers to the distance between training throws and the solution space of each of the eventual testing positions.

@@

@@

### Computing Similarity

@@

### Model Definition

@@

## Hit The Wall Task

@@

<br><br><br>
- 156 participants included in final analysis
- Varied group trains from 3 "velocity bands", constant group from 1
:::
::::

@@

## Project 2 - ALM

![](assets/fig-alm-diagram-1-1.png)

@@

## ALM

## EXAM

## Project 2 - Model Fitting Procedure {.r-fit-text}

### Approximate Bayesian Computation

@@

- simulation based approach
- approximate likelihood
- uncertainty in parameter values
- full distribution of plausible model predictions for each participant

@@

@@

## Modelling Results

![](assets/fig-ee-e1-1.png)

@@

>>>>

---
title: "The Role of Variability in Learning Generalization: A Computational Modeling Approach"
page-layout: full
aliases:
  - /slides.html
css: /Assets/Style/slides.css
title-slide-attributes:
  data-background-image: /Presentation/assets/sol_blank.png
  data-background-size: 300px #contain
  data-background-opacity: "0.15"
  data-background-position: "0% 0%"
format-links: [clean-revealjs, pptx,gfm]
format: 
  html: 
    subtitle: "Dissertation Defense"
  clean-revealjs:
    #theme: moon #white
    #output-ext: revealjs.html
    # template-partials:
    #   - assets/title-slide.html
    output-file: slides_revealjs
    date: 05/28/2024
    slideNumber: true
    hash-type: number # Use slide numbers in the URL hash
    center: true
    scrollable: true
    chalkboard:
      theme: whiteboard
  pptx:
    incremental: false
  hugo-md:
    include: false
    html-math-method: mathjax
    output-file: slides_hugo.md
  gfm:
    output: false
    output-file: slides_gfm.md
  # beamer:
  #   incremental: false
  #   colortheme: lily 
author:
  - name: Thomas Gorman
    url: https://tegorman13.github.io
    orcid: 0000-0001-5366-5442
    affiliations:
      - name: Indiana University
        department: Psychological & Brain Sciences, Cognitive Science
# params:
#   start_slide: 
#engine: knitr
---

@@

## Issues with previous research {.smaller}

@@

:::: columns
::: {.column}
![](assets/kerr_booth.png)
:::
::: {.column}
- Bean bag tossing task
- Constant and varied conditions train from distinct positions
- Both groups are tested from the position where the constant group trained
- Impressive demonstration of varied training outperforming constant training 
:::
::::
::: footer
@kerrSpecificVariedPractice1978
:::

@@

- Instances refers to individual throws (x and y velocity)
- Similarity refers to the distance between training throws and the solution space of each of the eventual testing positions.

@@

## Project 1 Computational Model {.smaller}

@@

#### Computing Similarity

@@

#### Model Definition

@@

## Hit The Wall Task {.smaller}

@@

- 156 participants included in final analysis
- Varied group trains from 3 "velocity bands", constant group from 1
- Both groups complete same total number of training trials
:::
::::

@@

## The Associative Learning Model ALM {.smaller}

::: {.column width="70%"}
![](assets/fig-alm-diagram-1-1.png)
:::
::: {.column width="30%"}
- Two layer network - adapted from ALCOVE (@kruschkeALCOVEExemplarbasedConnectionist1992)
- Input layer node for each stimulus, output node for each response 
- Input nodes activate as a function of their Gaussian similarity to the stimulus
- Weights udpated via delta rule - error driven learning
- Provides good account of human learning data, and interpolation performance, but struggles with extrapolation [@deloshExtrapolationSineQua1997]
:::
::::
------------------------------------------------------------------------

## Extrapolation - Association Model (EXAM) {.smaller}

- Extension to ALM to account for human extrapolation behavior 
- When a novel stimulus is presented, EXAM assumes the nearest 2-3 prior examples are retrieved, and used to compute a slope
- The slope is used to adjust the ALM response

|                    | **EXAM Response Generation**                                                                                     |                                                                                                               |
|--------------------|-----------------------------------------------------------------------------------------------------------------|---------------------------------------------------------------------------------------------------------------|
| Slope Computation  | $S = \frac{m(X_{1})-m(X_{2})}{X_{1}-X_{2}}$                                                                     | Slope value, $S$ computed from nearest training instances                                                     |
| Response           | $E[Y|X_i] = m(X_i) + S \cdot [X - X_i]$                                                                         | Final EXAM response is the ALM response for the nearest training stimulus, $m(X_i)$, adjusted by local slope $S$. |

@@

## Project 2 - Model Fitting Procedure {.smaller}

### Approximate Bayesian Computation (ABC)

@@

- simulation based parameter estimation [@turnerTutorialApproximateBayesian2012; @kangasraasioParameterInferenceComputational2019]
- Useful for models with unknown likelihood functions (e.g. many neural network and drift diffusion models)
- full distribution of plausible model predictions for each participant

@@

## Project 2 - Model Fitting Procedure {.smaller}

### Model Fitting Approach

- Two parameters for both ALM and EXAM
  - generalization parameter: $c$
  - learning rate parameter: $lr$

- ALM and EXAM fit separately to each individual participant
- Rejection based ABC used to obtain 200 posterior samples, per participant, per model 
  - i.e. 200 plausible values of $c$ and $lr$

- All models fit with three different approaches
    1) Fit to only the training data
    2) Fit to both training and testing data
    3) Fit to only testing data

@@

## Modelling Results {.smaller}

#### Posterior Predictive Distribution

![](assets/post_pred_e1.png)

::: notes
 Empirical data and Model predictions for mean velocity across target bands. Fitting methods (Test Only, Test & Train, Train Only) - are separated across rows, and Training Condition (Constant vs. Varied) are separated by columns. Each facet contains the predictions of ALM and EXAM, alongside the observed data.
:::

## Model Comparison - Experiment 1

![](assets/model_reg_e1.png)

::: notes
Figure 21: A-C) Conditional effects of Model (ALM vs EXAM) and Condition (Constant vs. Varied). Lower values on the y axis indicate better model fit. D) Specific contrasts of model performance comparing 1) EXAM fits between constant and varied training; 2) ALM vs. EXAM for the varied group; 3) ALM fits between constant and varied. Negative error differences indicate that the term on the left side (e.g., EXAM Constant) tended to have smaller model residuals.
:::

## Model Comparison - Experiment 2 and 3

![](assets/fig-e2_e3_ae-1.png)

::: notes
Conditional effects of Model (ALM vs EXAM) and Condition (Constant vs. Varied) on Model Error for Experiment 2 and 3 data. Experiment 3 also includes a control for the order of training vs. testing bands (original order vs. reverse order).
:::

## Modelling Results {.smaller}

### Best fitting models per participant

![](assets/best_models.png)

::: notes
Difference in model errors for each participant, with models fit to both train and test data. Positive values favor EXAM, while negative values favor ALM
:::

@@

## ALM {.scrollable .unnumbered .unlisted}

## EXAM {.scrollable .unnumbered .unlisted}
